---
title: "Low-Mg Carbonate XRD data analysis, 0-100% range calcite"
author: "John M. Jaeger"
date: "May 24, 2017"
output:
  word_document: default
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

##  code description

This  code is designed to import and analyze X-ray diffraction data collected at the University of Florida on a Rigaku Ultima IV diffractometer. The data are from carbonate mineral standards created to calibrate the diffractometer to provide quantitative measures of low- and high-Mg calcite. The general approach follows: 
Sepulcre, S., Durand, N., and Bard, E., 2009, Global and Planetary Change: Global And Planetary Change, v. 66, p. 1–9, doi: 10.1016/j.gloplacha.2008.07.008.

This code analyzes a full range of calcite/aragonite mixtures, with assumed values used for 100% calcite and 100% aragonite mixtures. These assumed values are used to better constrain polynomial fit.

Code generated by:
John M. Jaeger
Associate Professor
241 Williamson Hall
P.O. Box 112120
Dept. of Geological Sciences
University of Florida
Gainesville FL 32611-2120, USA
(352) 846-1381
ORCID ID# orcid.org/0000-0003-0248-489X
http://people.clas.ufl.edu/jmjaeger/

### Load packages
```{r load_packages, message=FALSE, results='hide'}
# if necessary uncomment and install packages.
# install.packages("dplyr")
# install.packages("knitr")
library(dplyr)
require(knitr)
library(ezknitr)
library(ezknitr)
library(ggplot2)
library(splines)
#library(propagate)

```


Import the data:
```{r load_data}
# load the End-member Dataset and give the dataset a shorter name

XRD.data <- read.table("../raw_data/UF_LMC_data_20170515.csv",header=T,sep=",") 
XRDdata <- tbl_df(XRD.data)

#testdata<-select(XRDdata, Sample, AreaR, LMC)
#area.ratio <- testdata$AreaR
#LMC.conc <- testdata$LMC
```
Calculate peak area ratios:
```{r peak area ratios}
area.ratio <- XRDdata$LMC_area/(XRDdata$LMC_area+XRDdata$Aragonite_area)
LMC.conc <- XRDdata$LMCconc
```

Develop polynomial calibration model for low-Mg calcite:
```{r lo_Mg ploynomial model}
model.fit3<- lm(LMC.conc~ poly(area.ratio,3))
summary(model.fit3)

# Confidence intervales for model parameters
confint(model.fit3, level=0.95)
```


Plot data and model:
```{r lo_Mg ploynomial model plot}

# Plot of fitted vs residuals
# No clear pattern should show in the residual plot if the model is a good fit
plot(fitted(model.fit3),residuals(model.fit3))


# Predicted values and confidence intervals
polypredicted.intervals <- predict(model.fit3,data.frame(x=area.ratio),interval='prediction',
                               level=0.95)

# Add lines to the existing plot
plot(area.ratio,LMC.conc,type='p',col='navy',main='Nonlinear relationship',lwd=1)

lines(area.ratio,polypredicted.intervals[,1],col='green',lwd=3)
lines(area.ratio,polypredicted.intervals[,2],col='black',lwd=1)
lines(area.ratio,polypredicted.intervals[,3],col='black',lwd=1)

# Add a legend
legend("topleft",c("Observ.","Predicted"), 
       col=c("deepskyblue4","green"), lwd=3)
```

The next step is to develop uncertainty in % LMC for a given area ratio, third-order polynomial. 
```{r predict poly3 value}
# Predicted values and confidence intervals
testvalue = data.frame(area.ratio=0.1)
predicted.value <- predict(model.fit3,testvalue,interval='prediction', level=0.95)
predicted.value
```





Now a spline model:
# https://www.r-bloggers.com/splines-opening-the-black-box/
# http://stackoverflow.com/questions/15837763/b-spline-confusion
# https://www.rdocumentation.org/packages/splines2/versions/0.2.4/topics/predict

```{r spline_fit}
library(splines)
spline1.pred <- lm(LMC.conc ~ bs(area.ratio, df=4))#df=4 means knots at end points and quantiles of X
summary(spline1.pred)
plot(area.ratio,LMC.conc)
u=seq(min(area.ratio),max(area.ratio),by=.05)
B=data.frame(area.ratio=u)
splinepredicted.interval=predict(spline1.pred,newdata=B,interval ="prediction",level=0.95)
lines(u,splinepredicted.interval[,1],lwd=2,col="green")
lines(u,splinepredicted.interval[,2],col='red',lwd=1)
lines(u,splinepredicted.interval[,3],col='red',lwd=1)

# Add a legend
legend("topleft",c("Observ.","Predicted"), 
       col=c("black","green"), lwd=3)
```
The last step is to develop uncertainty in % LMC for a given area ratio, spline model.
```{r predict spline value}
# Predicted values and confidence intervals
predicted.splinevalue <- predict(spline1.pred,testvalue,interval='prediction', level=0.95)
predicted.splinevalue
```
Export fits and prediction intervals:
```{r export results}
polyout <- cbind(area.ratio,polypredicted.intervals)
splineout <- cbind(u,splinepredicted.interval)
write.csv(polyout, file="../reports/UF_LMC_polyfit.csv")
write.csv(splineout, file="../reports/UF_LMC_splinefit.csv")


```

