---
title: "High-Mg Carbonate XRD data analysis with Xd,Xq; 0-5% calcite"
author: "John M. Jaeger"
date: "July 10, 2017"
output:
  word_document: default
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

##  Code Description

This  code is designed to import and analyze X-ray diffraction data collected at the University of Florida on a Rigaku Ultima IV diffractometer. The data are from carbonate mineral standards created to calibrate the diffractometer to provide quantitative measures of low- and high-Mg calcite. 

The general calibration approach follows: 
Sepulcre, S., Durand, N., and Bard, E., 2009, Global and Planetary Change: Global And Planetary Change, v. 66, p. 1–9, doi: 10.1016/j.gloplacha.2008.07.008.

Establishing detection and quantificiation limits follows:
Evard, H., Kruve, A., and Leito, I., 2016, Tutorial on estimating the limit of detection using LC-MS analysis, part I: Theoretical review: Analytica Chimica Acta, v. 942, p. 23–39, doi: 10.1016/j.aca.2016.08.043.




Code generated by:
John M. Jaeger
Associate Professor
241 Williamson Hall
P.O. Box 112120
Dept. of Geological Sciences
University of Florida
Gainesville FL 32611-2120, USA
(352) 846-1381
ORCID ID# orcid.org/0000-0003-0248-489X
http://people.clas.ufl.edu/jmjaeger/

## Load packages
```{r load_packages, message=FALSE, results='hide'}
# if necessary uncomment and install packages.
# install.packages("dplyr")
# install.packages("knitr")
library(dplyr)
require(knitr)
library(ezknitr)
library(chemCal)
library(reshape2)
library(RColorBrewer)
```

## Install functions
```{r function set up}
rnorm2 <- function(n,mean,sd) { mean+sd*scale(rnorm(n)) }
```


## Import the data
### Load peak area values for calibration standards
Low-Mg calcite peak at 29.4° 2-theta; Aragonite peak at 26.2° 2-theta (copper anode)
```{r load_data}
# load the End-member Dataset and give the dataset a shorter name

XRD.data <- read.table("../raw_data/UF_HMC_data_20170515.csv",header=T,sep=",") 
XRDdata <- tbl_df(XRD.data[1:18,])
XRDdata <- as.data.frame(XRDdata)
n = 3 #number of replicates of each calibration sample

```


### Establish ratios for aragonite blanks
We use peak areas for aragonite blank samples in the regression model (See Edvard et al. 2016a). Net peak areas are determined by subtracting an average background, which is the mean of eight measurements on either side of peak. First three blank samples are used here, but user can choose from any 24 total blank measurements.
```{r blank data, message=FALSE, results='hide'}
#Load aragonite blank data
aragonite.data <- read.table("../raw_data/UF_pure_AC_aragonite_XRD.csv",header=T,sep=",") 
aragonitedata <- tbl_df(aragonite.data)

# HMC peak region spans from 29.20-30.34° two-theta (Cu) based on 0.3% HMC standard
HMC_region <- aragonitedata[371:485,]
Cbkgdleft <- aragonitedata[363:370,]
Cbkgdright <- aragonitedata[486:493,]
Cbkgd.three <- rbind(Cbkgdleft[,2:4],Cbkgdright[,2:4])
Cmelt.bkgd <- melt(Cbkgd.three)
Cmeanbkgd <- mean(Cmelt.bkgd[,2])
HMC.net <- HMC_region[,2:4]-Cmeanbkgd
HMC.netPA <- colSums(HMC.net)

# Aragonite peak region spans from 25.58-26.60° two-theta (Cu) 
Arag_region <- aragonitedata[9:111,]
Abkgdleft <- aragonitedata[1:8,]
Abkgdright <- aragonitedata[112:119,]
Abkgd.three <- rbind(Abkgdleft[,2:4],Abkgdright[,2:4])
Amelt.bkgd <- melt(Abkgd.three)
Ameanbkgd <- mean(Amelt.bkgd[,2])
Arag.net <- Arag_region[,2:4]-Ameanbkgd
Arag.netPA <- colSums(Arag.net)
blank <- c(0,0,0)
HMC.blank <- rbind(blank,HMC.netPA,Arag.netPA)
HMCblank <- t(HMC.blank)
colnames(HMCblank) <- c("HMCconc", "HMC_area","Aragonite_area")

XRDdata.all <- rbind(HMCblank,XRD.data)#combine blank data with calibration data
XRDdata.HMC <- XRDdata.all[1:21,] # use only data from 0-5% HMC
calib.levels <- as.double(nrow(XRDdata.HMC))/n #number of calibration concentration levels
```

### convert peak areas to peak area ratios
```{r convert to peak area}
AreaRatio <- XRDdata.HMC$HMC_area/(XRDdata.HMC$HMC_area+XRDdata.HMC$Aragonite_area)
HMCconc <- XRDdata.HMC$HMCconc
HMC.lodata <- data.frame(HMCconc,AreaRatio)
HMClo <- tbl_df(HMC.lodata)
```

## Create ordinary least squares (OLS) regression model
We first develop an OLS linear calibration model for low-Mg calcite. We create a residual plot to examine for homoscedasticity and whether a linear model is an appropraite fit (see Edavrd et al., 2016 for rationale of these inital tests).
```{r HMC linear model}
HMC.lm = lm(AreaRatio~ HMCconc, data=HMClo)
vcov(HMC.lm)
summary(HMC.lm)
Bcalc = HMC.lm$coefficient[[2]]
Acalc = HMC.lm$coefficient[[1]]
Residual = AreaRatio - Bcalc * HMCconc - Acalc
# Data is plotted
plot(HMCconc, Residual)
```
Residuals show no trend and cluster around zero, so linear model is appropriate. Variance in residuals do  have distinctive "funnel" shape, so data are considered heteroscedastic. OLS model is not approprate, so WLS is preferred. Results using an OLS approach are presented at end for comparision with WLS results.


## Create weighted least squares (WLS) regression model 
We determine WLS calibration regression according to methods and R code of:
Evard, H., Kruve, A., and Leito, I., 2016, Tutorial on estimating the limit of detection using LC-MS analysis, part I: Theoretical review: Analytica Chimica Acta, v. 942, p. 23–39, doi: 10.1016/j.aca.2016.08.043.

```{r WLS model}

signal.data <- as.matrix(AreaRatio)
signal.data1 <- matrix(nrow = calib.levels, ncol = n)
conc.data1 <- matrix(nrow = calib.levels, ncol = 1)
for(u in 1:calib.levels){
    signal.data1[u,1] <- signal.data[3*u-2]
}
for(u in 1:calib.levels){
    signal.data1[u,2] <- signal.data[3*u-1]
}
for(u in 1:calib.levels){
    signal.data1[u,3] <- signal.data[3*u]
}
for(u in 1:calib.levels){
    conc.data1[u] <- HMCconc[3*u-2]

}

Cin <- unlist(conc.data1)
Y <- c(signal.data1[, 1:n])
C <- rep(Cin, n)

# Calculating weights for calibration samples
  SdRow <- NULL
  sdvalm <- NULL
  mean.PAR <- NULL
  for(k in 1:nrow(signal.data1)){
    SdRow[k] <- sd(signal.data1[k, ])^-2
    sdvalm[k] <- sd(signal.data1[k, ])
    mean.PAR[k] <- mean(signal.data1[k, ])
    }
  Nsamp <- nrow(signal.data1) * ncol(signal.data1)
  sumCal <- sum(SdRow)
  sumvar <- (sumCal) / (Nsamp) 
  w <- NULL
  #weights are normalized to make unitless; see Edvard et al. 2016 for justification
  for(v in 1:nrow(signal.data1)){
    w[v] <- sd(signal.data1[v, ])^-2 / sumvar
  }
  W <- rep(w, n)#weights for WLS model
  

  WLS <- lm(Y ~ C, weights = W)
  bw_calc = lm(Y ~ C, weights = W)$coeff[2]
  aw_calc = lm(Y ~ C, weights = W)$coeff[1]
  
  sigmodel <- bw_calc*Cin+aw_calc
  plot(C,Y, ann=FALSE)
  abline(aw_calc, bw_calc,col = "blue")# WLS fit
  abline(Acalc,Bcalc, col="red")#OLS fit
  title(xlab="HMC Concentration (mass%)")
  title(ylab="Peak Area Ratio")
  legend (0, max(Y), c("OLS model fit","WLS model fit"),col=c("red", "blue"),lty=c(1,1))
```
The slope and intercept of the two regression lines are very slightly different, which will affect the resulting limit calculations that depend on these values. See text for further discussion.

## Establish limits of detection and quantification
We determine the limits of decision, detection, and quantification (Yc, Xd, Xq, respectively) from calibration regression according to:
Burdge, J.R., MacTaggart, D.L., and Farwell, S.O., 1999, Realistic detection limits from confidence bands: J Chem Educ., v.76(3), p. 434-439  
and
Zorn, M.E., Gibbons, R.D., and Sonzogni, W.C., 1997, Weighted least-squares approach to calculating limits of detection and quantification by modeling variability as a function of concentration: Analytical chemistry, v. 69, p. 3069–3075.

The decision limit (Lc or Yc as used here) is the peak area ratio above which the signal is deemed at a 95% probabilty to rise above noise. The intersection of this y-value with the regression line is the Xc, or critical level in the concentration domain; however, it is suggested by Edvard et al. (2016) that the crtiical level be expressed only in the signal domain. The limit of detection (Xd here but often refered to as LoD) is the detection limit in concentration units (Yd here in signal units, sometimes denoted as Ld).
```{r WLS limits}
# Equations for Yc and Xd used here are from Burdge, J.R., MacTaggart, D.L., and Farwell, S.O., 1999, Realistic detection limits from confidence bands: J Chem Educ., v.76(3), p. 434-439  

#If no blank replicates are measured, it is necessary to create a model of signal st deviation as a function of concentration, which is then used to model weights for intercept and unknown values
stval.lm <- lm(sdvalm~poly(Cin,2,raw=TRUE)) #use if necessary to model St dev in calibration samples to assign weights; See Zorn et al. (1997)
p <- 0 #parameters in model of st dev versus concentration, which is zero for this because we use measured values applied to ranges closest in % HMC. Use 2 for linear model or 3 for 2nd order polynomial; See Zorn et al. (1997)
t.quantiles <- qt(c(.05, .95), Nsamp-p-2)# degrees of freedom in regression model
tval <- t.quantiles[2]# degrees of freedom in regression model
m <- 1# number of replicate measurements of unknown, usually 1 for establishing limits
#See Burge et al. (1999) for explantion of variables below:
xbarw <- sum(C*W)/sum(W)
Sxxw <- sum(W*C^2)-((sum(C*W))^2/sum(W))
Syyw <- sum(W*Y^2)-((sum(Y*W))^2/sum(W))
Sxyw <- sum(W*Y*C)-(sum(W*C)*sum(W*Y))/sum(W)
Swr <- sqrt((Syyw-bw_calc*Sxyw)/(Nsamp-2))
SwrN <- Swr/sqrt(sum(W)/Nsamp)
gw <- ((tval*Swr)^2)/(bw_calc^2*Sxxw)

Q <- sqrt((1/(m*W[1]))+(1/sum(W))+xbarw^2/Sxxw)
A <- ((tval*Swr*Q)/bw_calc)-(xbarw*gw)
B <- xbarw*gw*(xbarw*gw-(2*tval*Swr*Q/bw_calc))
Cv <- 1/sum(W)+xbarw^2/Sxxw+gw/(m*W[1])
Yc.wls <- aw_calc+tval*Swr*(sqrt(1/(m*W[1])+(1/sum(W))+(xbarw^2/Sxxw)))# decision limit in signal units
Xd.wls <- (A+(sqrt(B+(gw*Sxxw*(((1-gw)/(m*W[2]))+Cv)))))/(1-gw) # limit of detection in concentration units. We do not iterate to find precise weight at Xd as recommended by Burdge et al. and simply assume the weight at the calibration point closest to estimate of Xd (i.e., W[2] here)

# We can also esitmate limits of quantification
Yq.wls <- 10*(sqrt(1/SdRow[2]))+aw_calc #limit of quantification in signal units; From Zorn et al. (1997), eqn. 23
Xq.wls <- (Yq.wls-aw_calc)/bw_calc #limit of quantification in concentration units; From Zorn et al. (1997), eqn. 24


cat("WLS limit of detection (%HMC): ", round(Xd.wls,digits = 1),"   ")
cat("WLS limit of quantification (%HMC): ", round(Xq.wls,digits = 1),"   ")

```


## Establish prediction bands for WLS calibration regression
```{r prediction bands for WLS}
wls.predval <- matrix(nrow = calib.levels, ncol = m)

HMC.model <- seq(0, 5, by=0.01)# set up a range of HMC values spanning limits at 0.01% HMC intervals
PAR.model <- bw_calc*HMC.model+aw_calc #model peak area ratio (PAR) as a function of HMC.model
weight.modin <- cbind(Cin,w) # link calibration concentration measurement weights to concentrations

#instead of modeling weights as a polynomial function of %HMC, we choose to set weights based on measured values. Weights for HMC.model values are assigned based on nearest measured concentration replicates.
j <- length(PAR.model)
wls.predvaltest <- matrix(nrow = j, ncol = 1)
for(jv in 1:j){
  lookvalue <- HMC.model[jv]
  weightx <- which(abs(Cin-lookvalue)==min(abs(Cin-lookvalue)))
  wls.predvaltest[jv,1] <- weight.modin[weightx[1],2]
}
Inverse.HMCdata <- cbind(HMC.model,PAR.model,wls.predvaltest)
wls.predval2 <- matrix(nrow = j, ncol = m)

#We develop 95% prediction band equation from Zorn et al., equation 13
for(pv in 1:j){
    wls.predval2[pv,1] <-tval*Swr*(sqrt(1/(m*wls.predvaltest[pv,1])+(1/sum(W))+((Inverse.HMCdata[pv,1]-xbarw)^2/Sxxw)))
}
predband.uM <- PAR.model+wls.predval2
predband.lM <- PAR.model-wls.predval2

Inverse.HMCdata <- cbind(HMC.model,PAR.model,wls.predvaltest,wls.predval2,predband.uM,predband.lM)
colnames(Inverse.HMCdata) <- c("HMC.model", "PAR.model","weight.model","uncert.model","upper.uncert","lower.uncert")
cols<-brewer.pal(n=8,name="Dark2")

plot(C,Y, ann=FALSE)
lines(Inverse.HMCdata[,1],Inverse.HMCdata[,2], col = cols[8])
lines(Inverse.HMCdata[,1],Inverse.HMCdata[,5], col = cols[2], lty=1)
lines(Inverse.HMCdata[,1],Inverse.HMCdata[,6], col = cols[2], lty=1)
#segments(0, Yc.wls, HMC.xupperC, unknown.PAR,col =cols[3],lty=1,lwd=2)
title(xlab="HMC Concentration (mass%)")
title(ylab="Peak Area Ratio")
legend(0, max(Y), c("calib. data","model fit","prediction band"),col=c("black", cols[8],cols[2]), lty=c(1,1,1))

```

## Create an inverse model for establishing %HMC from a peak area ratio value
We create a model of fit and uncertainity to perform inverse modeling. We use the intercept of the P.A.R. with the regression line to determine the most likely %HMC value. We use the intercept of the P.A.R. with the prediction bands from the calibration curve to establish the 95% predition limits on the % HMC.
```{r inverse model from sample}

unknown.PAR <- 0.04 # value chosen to compare with model fits of Sepulcure et al. (2009) and Smodej et al. (20XX)

if(unknown.PAR > max(PAR.model)) {
      stop("Your unknown sample value is outside calibration range")
    }
HMC.x <- which(abs(PAR.model-unknown.PAR)==min(abs(PAR.model-unknown.PAR)))
HMC.inverse <- HMC.model[HMC.x ]# inverse fit of PAR that provides best fit to %HMC

unk.PARline <- rep(unknown.PAR,j)

#First, we find intersection of PAR value with upper prediction band
x1=unk.PARline
x2=Inverse.HMCdata[,5]
# Find points where x1 is above x2.
above<-x1>x2
# Points always intersect when above=TRUE, then FALSE or reverse
intersect.points<-which(diff(above)!=0)
# Find the slopes for each line segment.
x1.slopes<-x1[intersect.points+1]-x1[intersect.points]
x2.slopes<-x2[intersect.points+1]-x2[intersect.points]
# Find the intersection for each segment.
x.points<-intersect.points + ((x2[intersect.points] - x1[intersect.points]) / (x1.slopes-x2.slopes))
HMC.xlower <- HMC.model[x.points]
HMC.xlowerC <- HMC.xlower[which(abs(HMC.xlower-HMC.inverse)==min(abs(HMC.xlower-HMC.inverse)))]

#Next, we find intersection of PAR value with lower prediction band
x1=unk.PARline
x2=Inverse.HMCdata[,6]
# Find points where x1 is below x2.
below<-x1<x2
# Points always intersect when below=TRUE, then FALSE or reverse
intersect.points<-which(diff(below)!=0)
# Find the slopes for each line segment.
x1.slopes<-x1[intersect.points+1]-x1[intersect.points]
x2.slopes<-x2[intersect.points+1]-x2[intersect.points]
# Find the intersection for each segment.
x.points<-intersect.points + ((x2[intersect.points] - x1[intersect.points]) / (x1.slopes-x2.slopes))
HMC.xupper <- HMC.model[x.points]
HMC.xupperC <- HMC.xupper[which(abs(HMC.xupper-HMC.inverse)==min(abs(HMC.xupper-HMC.inverse)))]

if (length(HMC.xlowerC) == 0){
      stop("Your unknown sample value is outside prediction band")
    } 
if (length(HMC.xupperC) == 0){
      stop("Your unknown sample value is outside prediction band")
    } 



#display.brewer.pal(8,"Dark2")
#pdf("unknownHMCsample_UF.pdf", encoding = "MacRoman", width=7, height=5, useDingbats=FALSE)
plot(C,Y, ann=FALSE,ylim=c(0,0.5),xlim=c(0,5))
lines(Inverse.HMCdata[,1],Inverse.HMCdata[,2], col = cols[8])
lines(Inverse.HMCdata[,1],Inverse.HMCdata[,5], col = cols[2], lty=1)
lines(Inverse.HMCdata[,1],Inverse.HMCdata[,6], col = cols[2], lty=1)
segments(0, unknown.PAR, HMC.xupperC, unknown.PAR,col =cols[3],lty=1,lwd=2)
text(1, unknown.PAR-unknown.PAR/20,"Peak Area Ratio of test sample",cex=0.8)
segments(HMC.inverse, unknown.PAR, HMC.inverse, 0,col =cols[3],lty=1,lwd=2)
segments(HMC.xlowerC, unknown.PAR,HMC.xlowerC, 0,col =cols[3],lty=3,lwd=2)
segments(HMC.xupperC, unknown.PAR,HMC.xupperC, 0,col =cols[3],lty=3,lwd=2)
title(xlab="HMC Concentration (mass%)")
title(ylab="Peak Area Ratio")
legend(0,0.4, c("calib. data","model fit","prediction band","sample data"),col=c("black", cols[8],cols[2],cols[3]), lty=c(1,1,1,1))
#dev.off()

```

Provide the user with best fit and uncertainty for unknown sample from WLS regression model
```{r inverse model results from sample}
cat("WLS best fit, mass %HMC: ", round(HMC.inverse,digits = 1), "   ")
cat("WLS lower %HMC (95% CI): ", round(HMC.xlowerC,digits = 1),"   ")
cat("WLS upper %HMC (95% CI): ", round(HMC.xupperC,digits = 1),"   ")
```

##  Alternative ordinary least squares calibration approach
It is possible to determine Yc,Xd,and Xq from OLS calibration regression if data are considered homoscedatic according to:
Evard, H., Kruve, A., and Leito, I., 2016, Tutorial on estimating the limit of detection using LC-MS analysis, part I: Theoretical review: Analytica Chimica Acta, v. 942, p. 23–39, doi: 10.1016/j.aca.2016.08.043.
 
```{r limits from OLS regression}

Nsamp <- length(HMCconc)

#using method of Lavagnini & Magno (2006)
newdata = data.frame(HMCconc=0)
Yc.all <- predict(HMC.lm, newdata, interval="predict", level = 0.95, df=Nsamp-n)
Yc <- Yc.all[3]
Ld_data <- data.frame(AreaRatio=Yc)
ivtest=inverse.predict(HMC.lm,Yc,alpha=0.05)
Xc <- ivtest$Prediction
inverse.HMC.lm <- lm(HMCconc~AreaRatio, data=HMClo)
Xd.all <- predict(inverse.HMC.lm, Ld_data, interval="predict", level = 0.95, df=Nsamp-n)
Xd <- Xd.all[3]

#using method of Miller & Miller(1988) and Edvard et al. (2016)        
# Output parameters:
Bcalc <- HMC.lm$coefficient[[2]] # estimated slope
Syx <- ( sum(resid(HMC.lm)^2 ) / (Nsamp - 2) )^0.5 # Standard deviation of residuals
Acalc <-  HMC.lm$coefficient[[1]] # estimated intercept
Aerror  <-  summary(HMC.lm)$coef[1,2] # estimated standard deviation of intercept
CCaC <-  (1.645 * Syx)/Bcalc #also known as Yc
CCaS <- (1.645 * Syx)#also known as Lc or Xc
Xd.ols <- 3.3*Syx/Bcalc #CCb, also limit of determination in concentration, also known as LoD  
Xq.ols <- 10*Syx/Bcalc #Limit of quantification in concentration 
Xd.int <- 3.3* Aerror/Bcalc #limit of determination based only on varince at intercept
cat("OLS Decision value (peak area ratio): ", round(CCaS,digits = 3), "   ")
cat("OLS Limit of detection (%HMC): ", round(Xd.ols,digits = 1),"   ")
cat("OLS Limit of quantification (%HMC): ", round(Xq.ols,digits = 1),"   ")

```

The ordinary least squares decision limit (threshold) (CCa) based on peak area ratio is 0.024.
The ordinary least squares limit of determination in concentration units is 1.1% HMC
The ordinary least squares limit of quantification in concentration units is 3.3% HMC



We establish peak area ratio values from calibration associated with crtical 3% by mass low-Mg calcite cutoff, which is within of limit of quantification:
```{r peak_area_limits}
newdata = data.frame(HMCconc=3)
predict(HMC.lm, newdata, interval="predict") 
predict(HMC.lm, newdata, interval="confidence") 
calplot(HMC.lm)
```

Calculate  accuracy and precision for reverse model (95% CI) using a peak area ratio of 0.133 (predicited from linear model fit for a mass% of 3 for low-Mg calcite; see output above)
```{r inverse-model}
ivtest=inverse.predict(HMC.lm,0.133)
cat("Inverse fit (%HMC): ", round(ivtest$Prediction, digits = 1), "   ")
cat("Inverse confidence limits (%HMC): ", round(ivtest$`Confidence Limits`, digits = 1), "   ")
#ivtest$Prediction
#ivtest$`Confidence Limits`
```
Accuracy and precision for linear calibration model at crtical cutoff is 3.0 +/- 0.7% based on calibration data of April 2017.

Use linear model to predict low-Mg calcite concentration from peak area ratio of unknown sample (e.g., 0.100):
```{r unknown_samples}
ivtest=inverse.predict(HMC.lm,0.100)
cat("Inverse fit (%HMC): ", round(ivtest$Prediction, digits = 1), "   ")
cat("Inverse confidence limits (%HMC): ", round(ivtest$`Confidence Limits`, digits = 1), "   ")
```
Sample concentration is above limit of detection. Caution should be used aoplying absolute value because it is below limit of quantification.




